---
title: "Reproducible Research: Peer Assessment 1"
author: "Emi Harry"
date: "June 8, 2016"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, tidy = FALSE, fig.show = 'asis', fig.width=8, fig.height=6, fig.align = 'center', fig.path = 'figure/', fig.lp = "Figure.", fig.keep = 'all')
```

## Section 1
### Loading and preprocessing the data

Read dataset into R

```{r}
fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(fileurl, temp) 
unzip(temp, list = T)
dataset1 <- read.table(unzip(temp, "activity.csv"), header = T, sep = ",", stringsAsFactors = F)
unlink(temp)
```


## Section 2

Next, to find the total steps per day

- Load the data table package. 
- Convert to data table and convert the data column variables to dates.
- Remove the NA's

```{r}
library(data.table)
dt <- as.data.table(dataset1)
dt$date <- as.Date(dt$date)

# remove the NA's
dt2 <- na.omit(dt)
```

- Load the dply package and use it to group by date and create the subset.
- Summarise the steps each and save in a new dataset

```{r}
library(dplyr)
dtgroup <- group_by(dt2, date)
dtanalysis <- summarise(dtgroup, sum(steps))
```

Check the column names and reset it to apprioprate names

```{r}
names(dtanalysis)
setnames(dtanalysis, c("date", "sum(steps)"),
         c("date", "total_steps"))
```
Plot the histogram of the total daily steps

```{r}
plot(dtanalysis$total_steps, type = "h", lwd = 10,
     main = "Total Daily Steps", xlab = "Days", ylab = "Steps", col = 6)
abline(h = mean(dtanalysis$total_steps), col = 2)
abline(h = median(dtanalysis$total_steps), col = 3)

```
### What is mean total number of steps taken per day?

Finding the mean and median of steps derived from the total daily steps.

```{r}
firstmean <- mean(dtanalysis$total_steps)
firstmedian <- median(dtanalysis$total_steps)
```
The mean is `r firstmean` and the median is `r firstmedian`.

### Section 3

Group by time interval and find thier averages

```{r}
intvervalgroup <- group_by(dt2, interval)
dtinterval <- summarise(intvervalgroup, mean(steps))
```

### What is the average daily activity pattern?

Plot time series of average steps taken

```{r}
plot(dtinterval$interval, dtinterval$`mean(steps)`, main = "Time Series of Average Steps (5 interval)",
     type = "l", xlab = "5-minute Interval", ylab = "Average Steps")
```

Set the interval as row names and get the corresponding value for the max mean

```{r}
df <- as.data.frame(dtinterval)
row.names(df) <- dtinterval$interval
maxindex <- which.max(dtinterval$`mean(steps)`)
max5 <- rownames(df)[maxindex]
```

The 5-minute interval with the highest average steps is `r max5`.



## Section 4
### Imputing missing values
load the mice package and view the pattern of the missing data


```{r fig.height=5, fig.width=4}
library(mice)
md.pattern(dataset1)

library(VIM)
aggr_plot <- aggr(dataset1, col=c('magenta','gray'), numbers=TRUE, sortVars=TRUE, 
                  labels=names(dataset1), cex.axis=.7, gap=3, 
                  ylab=c("Histogram of missing data","Pattern"))
```

use the predictive mean matching method in imputing the missing data
```{r}
tempData <- mice(dataset1,m=5,maxit=10,meth='pmm',seed=500)

summary(tempData)
completedData <- complete(tempData,2)
```

To find the total steps per day, convert to data table and subset by date.
Then sum the steps in each subset. Use the dplyr package to do the grouping.

```{r}
completedData <- as.data.table(completedData)
comp <- group_by(completedData, date)
ctotal <- summarise(comp, sum(steps))
cmean <- mean(ctotal$`sum(steps)`)
cmedian <- median(ctotal$`sum(steps)`)
```

After inputing the missing values, the new mean is `r cmean` and the median is `r cmedian`.

There is a difference between the new mean and the old mean because when the first one was calculated, the NA rows were omitted but the is no difference in the median. Hence there were fewer observations to work with but by replacing the NA's using the predictive mean matching method, it the number of observations increased and the total daily averages silghtly, resulting in a lower mean.

plot the histogram of the total daily steps after NA values are added.

```{r}
plot(ctotal$`sum(steps)`, type = "h", lwd = 7,
     main = "Total Daily Steps", xlab = "Days", ylab = "Steps", col = 11)

```

## Section 5

View date as weekdays

```{r}
comp$date <- as.Date(comp$date)
comp$date <- weekdays(comp$date)
```

Create 2 subsets for weekday and weekend 

```{r}
dtweekday <- comp[grep("Monday|Tuesday|Wednesday|Thursday|Friday", date, ignore.case =T)]
wkdy <- group_by(dtweekday, interval)
wkdymean <- summarise(wkdy, mean(steps))

dtweekend <- comp[grep("Saturday|Sunday", date, ignore.case = T)]
wknd <- group_by(dtweekend, interval)
wkndmean <- summarise(wknd, mean(steps))
```

Add the day_type column to both datasets

```{r}
wkdymean <- data.table(wkdymean, day_type = "weekday")
wkndmean <- data.table(wkndmean, day_type = "weekend")
```

Row bind the 2 datasets to creat a new factor variable in the dataset with two levels - "weekday" and "weekend"

```{r}
dtgroup3 <- rbind(wkdymean, wkndmean)
```

Use the weekday and weekend datasets to plot the graphs for 5-minute intervals

```{r}
library(lattice)
wk <- xyplot(dtgroup3$`mean(steps)` ~ dtgroup3$interval | factor(dtgroup3$day_type), type = "l", main = "Average Steps per 5-minute Interval", ylab = "Number of Steps", xlab = "Interval", layout = c(1,2))
print(wk)
```

### Are there differences in activity patterns between weekdays and weekends?

From this panel plot, it is clear that the highest mean steps occurs on the weekdays but the weekend has the higher median steps.

